---
title: "StarterPack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{StarterPack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r lib}
library(R2Excel)
```

## Before start

Bellow play with some basic examples, check

```{r setup, eval = FALSE}
packageVersion("R2Excel")
utils::news(package = "R2Excel")
```

```r
dir.create("tmp", showWarnings = FALSE) # prepare the test folder ...
```

## Basic Data

We propose to play with prepared toy data

```r
str(modified_sleep) # see the data's structure used with str()
```

The main function of the package `R2Excel` is `save_excel_results` where you will provide the dataset you want to describe with several options.

(1) It can be a global description, without variable of stratification, so let `varstrat = NULL` by default.  

``` r
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp", "01-desc_tab.xlsx"),
  vars = c(
    "Population", "Income", "Illiteracy", "Life Exp", "Murder",
    "HS Grad", "Frost", "Area", "state.division", "state.region", "binary_test"
  ),
  varstrat = NULL,
  digits = 3,
  simplify = FALSE,
  light_contents = TRUE
)
```

(2) Otherwise, name the column of a factor to separate the statistics, for instance with `varstrat = election`.

```r
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp", "02-desc_tab_strat.xlsx"),
  vars = c(
    "Population", "Income", "Illiteracy", "Life Exp", "Murder",
    "state.division", "state.region", "binary_test"
  ),
  varstrat = "election",
  digits = 3,
  simplify = FALSE,
  light_contents = TRUE
)
```

(3) You may be want to remove redundant information about binary (0/1 or non/oui) variables, 
use `simplify` ; may be useful to see also `force_generate_1_when_0`.

``` r
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp", "03-desc_tab.xlsx"),
  vars = c(
    "Population", "Income", "Illiteracy", "Life Exp", "Murder",
    "yes_no_french_question", "all_count_zero"
  ),
  varstrat = "election",
  digits = 3,
  simplify = TRUE, 
  force_generate_1_when_0 = TRUE
)
```


Then, just in R and not especially saving your table in an Excel file, 
you can get quick and beautiful tables of descriptive statistics. 
See `compute_*` functions.

(4) Either for continuous variables : 

``` r
vars_wanted <- c(
  "Population", "Income", "Illiteracy", "Life Exp", "Murder",
  "HS Grad", "Frost", "Area"
)
varstrat_wanted <- "election"

res_conti_tabs <- compute_continuous_table(
  dataframe = modified_state,
  vars = vars_wanted,
   varstrat = varstrat_wanted,
   digits = 2
)
res_conti_tabs
compute_continuous_table(
  dataframe = modified_state,
  vars = vars_wanted[1],
  varstrat = varstrat_wanted,
  stats_choice = c("mean", "sd", "N"),
  digits = 4
)
```

(5) Nor for factorial variables : 

``` r
vars_wanted <- c("state.division", "state.region", "binary_test")
varstrat_wanted <- "election"
res_fact_tabs <- compute_factorial_table(
  dataframe = modified_state,
  vars = vars_wanted,
  varstrat = varstrat_wanted,
  digits = 2,
  simplify = FALSE
)
res_fact_tabs

# also possible to simply binary info
vars_wanted <- c("state.division", "state.region", "binary_test", "yes_no_french_question")
varstrat_wanted <- "election"
res_fact_simpl_tabs <- compute_factorial_table(
  dataframe = modified_state,
  vars = vars_wanted,
  varstrat = varstrat_wanted,
  digits = 2,
  simplify = TRUE
)
res_fact_simpl_tabs

# and general zero with or without simplification
vars_wanted <- c("yes_no_french_question", "all_count_zero")
varstrat_wanted <- "election"
res_fact_simpl0_tabs <- compute_factorial_table(
  dataframe = modified_state,
  vars = vars_wanted,
  varstrat = varstrat_wanted,
  digits = 2,
  simplify = FALSE,
  force_generate_1_when_0 = TRUE
)
res_fact_simpl0_tabs
```

To go further on statistics, you can test tables according your studied `varstat`. 
See `test_*` functions.

(6) Tests for qualitative variables

``` r
res_prop <- test_proportions(
  dataframe = modified_state[, c("state.division","state.region", "binary_test", "election")],
  vars = c("state.region", "state.division", "binary_test"),
  varstrat = "election",
  digits = 3
)
res_prop
``` 

(7) Tests for quantitative variables

``` r
res_means <- test_means(
  dataframe = modified_state[, c("Population", "Income", "election")],
  vars = c("Population", "Income"),
  varstrat = "election",
  digits = 3
)
res_means
```

(8) come with the estimated difference of means

``` r
estim_diff <- estimate_diff_mean(
  dataframe = modified_state[, c("Population", "Income", "election")],
  vars = c("Population", "Income"),
  varstrat = "election",
  digits = 2
)
estim_diff
```

(9) With `save_excel_results`, several new possibilities are available.  
You can now describe 2 varstat crossed (one as a subgroup of the second one) is the varstrat params is provided 
as "var1*var2" or as c("var1", "var2") :

```r 
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp", "09-crossed_desc.xlsx"),
  vars = c(
    "Population", "Income", "Illiteracy", "Life Exp", "Murder",
    "HS Grad", "Frost", "Area",
    "state.division", "state.region"
  ),
  varstrat = "election*binary_test",
  digits = 2
)

# without the Population_totale and NB detail (as sum) but with stat test inside varstrat1
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp", "09-crossed_desc_test.xlsx"),
  vars = c(
    "Population", "Income", "Illiteracy", "Life Exp", "Murder",
    "HS Grad", "Frost", "Area",
    "state.division", "state.region"
  ),
  varstrat = "election*binary_test",
  digits = 2, 
  global_summary = FALSE,
  crossed_varstrat_test = TRUE,
  detail_NB_mesure_sum = FALSE
)
``` 

(10) The option `varstrat` can now take a continuous variable : 

Based on `compute_continuous_table` (transposed) table for each qualitative levels and `compute_correlation_table` with it test for quantitative vars (correlation method can be "detect_auto" or classical known approachs)

```r
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp","10-conti_varstrat.xlsx"),
  vars = c(
    "Population", "Income", "Illiteracy", "Life Exp",
    "state.division", "state.region", "election"
  ),
  varstrat = "Area",
  digits = 2
)
```

```r
compute_correlation_table(
  dataframe = modified_state,
  vars = c("Population", "Income"),
  varstrat = "Area",
  method = "detect_auto",
  digits = 2
)
```

(11) The option `show_OR` add the univariate OR columns when the varstrat is a binary factor. 
It is based on `get_OR_univar` function.

```r
get_OR_univar(
   dataframe = modified_state,
   dependent_var = "binary_test",
   explanatory_vars = c(
     "Population", "Income", "Life Exp", "state.region",
     "yes_no_french_question", "all_count_zero"
   )
)
```

```r
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp","11-OR_tab.xlsx"),
  vars = c(
    "Population", "Income", "Life Exp", "state.region",
    "yes_no_french_question", "all_count_zero"
  ),
  varstrat = "binary_test",
  digits = 2, 
  simplify = FALSE,
  show_OR = TRUE
)
```

(12) The option `show_OR` and `simplify` can work together too.

```r
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp","12-OR_tab.xlsx"),
  vars = c(
    "Population", "Income", "Life Exp", "state.region",
    "yes_no_french_question", "all_count_zero"
  ),
  varstrat = "binary_test",
  digits = 2, 
  simplify = TRUE,
  show_OR = TRUE
)
```


(13) you might be interested in putting labels in front of your complicated variable names.


```r
save_excel_results(
  dataframe = modified_state,
  file = file.path("tmp", "13-labels.xlsx"),
  vars = c(
    "Population", "Illiteracy", "Income",
    "state.division", "state.region"
  ),
  varstrat = "binary_test",
  dico_mapping = data.frame(
    "Vars" = c("Population", "Income", "state.division", "state.region"), 
    "labels" = c("Population size (nb)", "Income (in dollars)", "State Division ???", "state region (cad point)")
  )
)
```

## Longitudinal (Paired) Data

If you have longitudinal measures (paired data), you should use "paired" family functions, 
for instance on toy data `modified_sleep`.

```r
str(modified_sleep)
```

(A) You can start by "compute" a "paired" table for continuous traits 

``` r
compute_paired_continuous_table_and_test(
  dataframe = modified_sleep,
  variable_interest = "extra",
  varstrat = "visites_2",
  patient_id = "ID2",
  digits = 2,
  global_summary = FALSE
)
compute_paired_continuous_table_and_test(
  dataframe = modified_sleep,
  variable_interest = "mesure3",
  varstrat = "visites_2",
  patient_id = "ID2",
  digits = 2,
  global_summary = FALSE
)
compute_paired_continuous_table_and_test(
  dataframe = modified_sleep,
  variable_interest = "extra",
  varstrat = "visites_5",
  patient_id = "ID",
  digits = 2,
  global_summary = FALSE
)
```


(B) or "compute" a "paired" table for factorial traits 

``` r
compute_paired_factorial_table_and_test(
  dataframe = modified_sleep,
  variable_interest = "fact1",
  varstrat = "visites_2",
  digits = 1,
  patient_id = "ID2",
  force_generate_1_when_0 = FALSE,
  keep_missing_line = TRUE
)
compute_paired_factorial_table_and_test(
  dataframe = modified_sleep,
  variable_interest = "fact1_na",
  varstrat = "visites_2",
  digits = 1,
  patient_id = "ID2",
  force_generate_1_when_0 = FALSE,
  keep_missing_line = TRUE
)
compute_paired_factorial_table_and_test(
  dataframe = modified_sleep,
  variable_interest = "fact3",
  varstrat = "visites_2",
  digits = 1,
  patient_id = "ID2",
  force_generate_1_when_0 = FALSE,
  keep_missing_line = TRUE
)
```

(C You can generalize the call of paired tables, for several variables of interest, saved in excel file. 

For instance with 2 visits

```r
save_excel_paired_results(
  dataframe = modified_sleep,
  vars = c("extra", "extra_with_missings", "mesure1", "mesure2", "mesure3"),
  varstrat = "visites_2",
  patient_id = "ID2",
  digits = 1,
  global_summary = TRUE,
  file = file.path("tmp", "C-desc_paired_data1.xlsx")
)
```

(D) For instance with 2 visits

```r
save_excel_paired_results(
  dataframe = modified_sleep,
  vars = c("extra", "extra_with_missings", "mesure1", "fact1", "fact1_na"),
  varstrat = c("visites_5"),
  patient_id = "ID",
  digits = 2,
  global_summary = TRUE,
  force_generate_1_when_0 = FALSE,  # for fact tab
  keep_missing_line = TRUE, # for fact tab
  file = file.path("tmp", "D-desc_paired_data3.xlsx")
)

# with stat test
save_excel_paired_results(
  dataframe = modified_sleep,
  vars = c("extra", "extra_with_missings", "mesure1", "fact1", "fact1_na"),
  varstrat = c("visites_5"),
  patient_id = "ID",
  digits = 2,
  global_summary = TRUE,
  force_generate_1_when_0 = FALSE,
  keep_missing_line = TRUE,
  test_more_2_levels = TRUE,
  file = file.path("tmp", "D-desc_paired_data_tested.xlsx")
)
```

(E) The option about "cross" 2 varstrats is also available in "paired" function : 

```r
save_excel_paired_results(
  dataframe = modified_sleep,
  vars = c("extra", "extra_with_missings", "mesure1", "fact1", "fact1_na"),
  varstrat = "visites_2*fact3",
  patient_id = "ID2",
  digits = 2,
  global_summary = TRUE,
  force_generate_1_when_0 = FALSE,  # for fact tab
  keep_missing_line = TRUE, # for fact tab
  file = file.path("tmp","E-nested_paired_desc.xlsx")
)
```

(F) understand `save_excel_paired_results_filtertest` : A specific table for repeated data may be wanted in two ways: to show rows with missing values to describe everything or to remove missing data to describe only tested data.  
Here is a snippet of code to do both :

```r
## description in detail with NA at each VISITE
R2Excel::save_excel_paired_results(
  dataframe = modified_sleep, 
  vars = c("extra_with_missings", "extra"),
  varstrat = "visites_2",
  patient_id = "ID2", 
  file = file.path("tmp", "F-missing_paired_desc.xlsx"),
  global_summary = FALSE,
  keep_missing_line = FALSE
)

## or only describe data tested (without missing values at each variable)
desc_data_tested <- data.table::rbindlist(l = lapply(
  X = c("extra_with_missings", "extra"),
  FUN = function(var_i) {
    
    tmp_i <- modified_sleep
    # get if with NA :
    detect_missing_id <- tmp_i[["ID2"]][which(is.na(tmp_i[[var_i]]))]
    # remove them
    tmp_i <- tmp_i[!ID2 %in% detect_missing_id, ]
    # or get id with 1 line ... (other kind of missing values)
    detect_missing_id <-local({
      tt <- as.data.frame(table(tmp_i[["ID2"]]))
      tt$Var1[tt$Freq == 1]
    })
    # remove them
    tmp_i <- tmp_i[!ID2 %in% detect_missing_id, ]
    
    # nrow(tmp_i)
    R2Excel::save_excel_paired_results(
      dataframe = tmp_i, 
      vars = var_i,
      varstrat = "visites_2",
      patient_id = "ID2", 
      file = file.path("tmp", paste0("F-Nomissing_paired_desc_", var_i, ".xlsx")),
      global_summary = FALSE,
      keep_missing_line = FALSE
      # metric_show = if wanted ? sepecifically for each var_i
    )
    # read the excel table
    tab <- readxl::read_excel(path = file.path("tmp", paste0("F-Nomissing_paired_desc_", var_i, ".xlsx")))
    unlink( # if want to clean temporary file
        file.path("tmp", paste0("F-Nomissing_paired_desc_", var_i, ".xlsx"))
    ) 
    return(tab)
  }
))
writexl::write_xlsx(
  x = list(
    "quantitative - visites_2" = desc_data_tested
  ),
  path = file.path("tmp", "F-Nomissing_paired_desc.xlsx")
)

## this last example is now include in a function meta function :
save_excel_paired_results_filtertest(
  dataframe = modified_sleep,
  vars = c("extra", "extra_with_missings", "group"),
  varstrat = "visites_2",
  patient_id = "ID2",
  file = file.path("tmp", "desc_paired_data_tested.xlsx")
)
```
